---
output:
  pdf_document: default
  html_document: default
---

<!-- # : Lane_County_Jail.rmd -->

<!-- # Author: @russl_corey -->

<!-- # Date: Mar 7, 2023 -->

<!-- # This program is free software: you can redistribute it and/or modify it under  -->

<!-- # the terms of the GNU General Public License as published by the Free Software  -->

<!-- # Foundation, either version 3 of the License, or (at your option) any later  -->

<!-- # version. -->

<!-- #  -->

<!-- # This program is distributed in the hope that it will be useful, but WITHOUT ANY  -->

<!-- # WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A  -->

<!-- # PARTICULAR PURPOSE. See the GNU General Public License for more details. -->

<!-- # You should have received a copy of the GNU General Public License along with  -->

<!-- # this program. If not, see <https://www.gnu.org/licenses/>.  -->

---
title: "Lane County Jail Analysis"
author: Russell_Corey
date: March 7, 2023
output: html_document
---

\newpage

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tufte)
```

```{r set working director and load data, echo=FALSE, message=FALSE, warning=FALSE}

setwd('/home/russell/Dropbox/DataAnalysis/Lane_County_Jail')

# Use the date that we downloaded booking data
study_date <-  as.Date("2023-03-09")

# Load charges from csv file
charges <- read_csv('data/inmate_list.csv', show_col_types = FALSE)

# Parse the code from the description
charges$v_code <- gsub("^(\\d+\\.?\\d*).*", "\\1", charges$`Violation:`)

# Get booking information from charges
inmates <- charges %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# Format dates
inmates$`Sched. Release:` <- as.Date(inmates$`Sched. Release:`, format = "%m/%d/%Y")
inmates$`Released:` <- as.Date(inmates$`Released:`, format = "%m/%d/%Y")


```

# Intro
This report aims to provide insight into the current detained population in the 
Lane County Jail on 2023-03-09. We will use a Python web scrapper to download the data from the 
jail's online inmate search, then format that data into a single table that we 
will then inspect and generate statistics about the population therein.

Source at [https://github.com/russl-corey/Lane_County_Jail](https://github.com/russl-corey/Lane_County_Jail)

# Data

Inmate data was retrieved from the Lane County Jail's booking search web portal:

[http://inmateinformation.lanecounty.org/Home/BookingSearchQuery](http://inmateinformation.lanecounty.org/Home/BookingSearchQuery)

on March 9th, 2023.

\newpage 

# Inmates

## Sex
```{r current inmates, echo=FALSE, warning=FALSE}


# filter out inmates without release dates
inmates <- inmates[is.na(inmates$`Released:`), ]

# Display total number of inmates in system
print(paste('Total bookings found:', length(unique(inmates$`full name`))))

# Create table with gender info
inmate_gender <- table(inmates$`Sex:`)
names(inmate_gender) = c('Female', 'Male')

# Display table 
inmate_gender %>%
  kable(caption='Sex Counts')

# Create a pie chart with labels and colors, and show percentages and counts
pie(inmate_gender, labels = paste0(names(inmate_gender), " (", inmate_gender, ")"), 
    col = c("#FF7F50", "#6495ED"), main = "Gender Distribution of Inmates", 
    cex = 0.8, clockwise = TRUE, density = 10)

# Add a legend with percentages and counts
legend("topright", c(paste0(names(inmate_gender), " (", round(100*inmate_gender/sum(inmate_gender), 1), "%, ", inmate_gender, ")")), cex = 0.8, bty = "n")


```
\newpage

## Age


```{r inmate age, echo=FALSE}
# Add a new column to indicate age group by decade
inmates$age_group <- cut(inmates$`Age:`, breaks = seq(0, 100, by = 10), include.lowest = TRUE)

# Count the number of inmates in each age group
counts <- table(inmates$age_group)

# Display table of counts
counts %>%
  kable(caption='Age Range Groupings')

# Create a bar chart with age group distribution
plot <- barplot(counts, main = "Age Group Distribution", xlab = "Age Group", ylab = "Count", col = rainbow(length(counts)))

# Add a labeled line at age 18
#abline(v = 18, col = "red")
#text(x = 18, y = max(counts), label = "18 years old", pos = 1)

```
\newpage

## Race


```{r inmate race, echo=FALSE, warning=FALSE}
# load necessary libraries
library(ggplot2)

# Count inmate race
inmate_counts <- inmates %>% 
  group_by(`Race:`) %>%
  summarize(count = n())

# Calc percentage
inmate_counts$percentage <- paste0(round(inmate_counts$count / sum(inmate_counts$count) * 100, 0), '%')

# Show table 
inmate_counts %>%
  kable(caption='Inmate Races')

# create the pie chart
ggplot(inmate_counts, aes(x="", y=count, fill=`Race:`)) +
  geom_bar(stat="identity", width=1) +
  coord_polar(theta = "y") +
  labs(title = "Inmate Race Ratios", fill = "Race", x = NULL, y = NULL) +
  theme_void() +
  theme(legend.position = c(1.0, 0.8))

```


\newpage

# Charges

## Measure 11

One of the metrics the sheriff has used is concerning inmates awaiting Measure 11 charges. 
Measure 11 defines minimum mandatory sentences for listed felonies. 


> "No pre-trial adults in custody being held on Measure 11 charges have been
> released due to capacity in nearly 10 years (since July 1, 2013)."
> `r tufte::quote_footer('--- Sheriff Cliff Harrold to Board of County Commissioners, on Jan 25, 2023')`

So let us inspect the porportion of inmates that are currently booked on Measure 11 charges.

```{r measure11 stats, echo=FALSE, warning=FALSE}

# Load list of measure 11 codes
measure11 <- read_csv('measure_11.csv', show_col_types = FALSE)

# encode TRUE/FALSE for measure 11 violations
charges$m11 <- FALSE
charges$m11[charges$v_code %in% measure11$code] <- TRUE

# Filter for only m11 charges then get distinct booking info
m11_booking <- charges %>%
  filter(m11 == TRUE) %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# add m11 column to inmate data frame
inmates$measure11 <- FALSE 
inmates$measure11[inmates$`Booking Number:` %in% m11_booking$`Booking Number:`] <- TRUE

# Count the number of inmates labeled as measure 11 and not
measure11_counts <- table(inmates$measure11)

# Create a pie chart with percentage labels
pie(measure11_counts, 
    labels = paste0(names(measure11_counts), ": ", 
                    round(100*measure11_counts/sum(measure11_counts), 2), "%"), 
    col = c("orange", "lightblue"),
    main = "Proportion of inmates with measure 11 charges")

# Add number annotations
legend("topright", legend = names(measure11_counts), 
       title = "Measure 11", 
       fill = c("orange", "lightblue"))

```




```{r measure11 inmates, echo=FALSE, warning=FALSE}

# Filter for only m11 charges then get distinct booking info
m11_booking <- charges %>%
  filter(m11 == TRUE) %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# Order and display inmates
m11_booking %>%
  arrange(`full name`) %>%
  kable(caption='Inmates with Measure 11 Charges')


charges %>%
  filter(m11 == TRUE) %>%
  group_by(`full name`, `v_code`) %>%
  ungroup() %>%
  select(`full name`, `Violation:`) %>%
  arrange(`full name`) %>%
  kable(caption='Inmates with Measure 11 charges')

```

\newpage

## Top Charges

```{r violations, echo=FALSE, warning=FALSE}

# Count all Violations
charge_counts <- charges %>%
  group_by(v_code) %>%
  summarize(count = n()) %>%
  arrange(desc(`count`))

# make list of codes to names
vcode_names <- charges %>%
  select(v_code, `Violation:`) %>%
   distinct(v_code, .keep_all = TRUE)

# merge violation names to violation counts
charge_counts <- merge(x=charge_counts, y=vcode_names,
                    by.x='v_code', by.y='v_code',
                    all.x=TRUE, all.y=FALSE)

# sort and display table of top violations
charge_counts %>%
  select(`Violation:`, count) %>%
  arrange(desc(`count`)) %>%
  head(20) %>%
  kable()


```


```{r inmates 2, echo = FALSE, warning=FALSE}

```

\newpage




# Appendix A


```{r A1, echo=FALSE}

```

```{r A2, echo = FALSE}


```



\newpage



# Appendix Z

source code available at <https://github.com/russl-corey/Lane_County_Jail>

This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details. You should have received a copy of the GNU General Public License along with this program. If not, see [\<https://www.gnu.org/licenses/\>](https://www.gnu.org/licenses/).
