---
output:
  html_document: default
  pdf_document: default
---

<!-- # : Lane_County_Jail.rmd -->

<!-- # Author: @russl_corey -->

<!-- # Date: Mar 7, 2023 -->

<!-- # This program is free software: you can redistribute it and/or modify it under  -->

<!-- # the terms of the GNU General Public License as published by the Free Software  -->

<!-- # Foundation, either version 3 of the License, or (at your option) any later  -->

<!-- # version. -->

<!-- #  -->

<!-- # This program is distributed in the hope that it will be useful, but WITHOUT ANY  -->

<!-- # WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A  -->

<!-- # PARTICULAR PURPOSE. See the GNU General Public License for more details. -->

<!-- # You should have received a copy of the GNU General Public License along with  -->

<!-- # this program. If not, see <https://www.gnu.org/licenses/>.  -->

---
title: "Lane County Jail Analysis"
author: Russell_Corey
date: March 7, 2023
output: html_document
---

\newpage

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tufte)
```

```{r set working director and load data, echo=FALSE, message=FALSE, warning=FALSE}

setwd('/home/russell/Dropbox/DataAnalysis/Lane_County_Jail')

# Use the date that we downloaded booking data
study_date <-  as.Date("2023-03-09")

# Load charges from csv file
charges <- read_csv('data/inmate_list.csv', show_col_types = FALSE)

# Parse the code from the description
charges$v_code <- gsub("^(\\d+\\.?\\d*).*", "\\1", charges$`Violation:`)

# Get booking information from charges
inmates <- charges %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# Format dates
inmates$`Sched. Release:` <- as.Date(inmates$`Sched. Release:`, format = "%m/%d/%Y")
inmates$`Released:` <- as.Date(inmates$`Released:`, format = "%m/%d/%Y")

charges <- charges[!is.na(charges$`full name`),]
```

# Intro
This report aims to provide insight into the current detained population in the 
Lane County Jail on 2023-03-09. We will use a Python web scrapper to download the data from the 
jail's online inmate search, then format that data into a single table that we 
will then inspect and generate statistics about the population therein.

Source at [https://github.com/russl-corey/Lane_County_Jail](https://github.com/russl-corey/Lane_County_Jail)

# Data

### Lane County Jail 
Inmate data was retrieved from the Lane County Jail's booking search web portal:

[http://inmateinformation.lanecounty.org/Home/BookingSearchQuery](http://inmateinformation.lanecounty.org/Home/BookingSearchQuery)

on March 9th, 2023.

### Oregon 2020 Census

The 2020 Oregon census data can be found at 

<https://www.pdx.edu/population-research/census-data-oregon>


\newpage 

# Inmates

## Sex
```{r current inmates, echo=FALSE, warning=FALSE}


# filter out inmates without release dates
inmates <- inmates[is.na(inmates$`Released:`), ]

# Display total number of inmates in system
print(paste('Total bookings found:', length(unique(inmates$`full name`))))

# Create table with gender info
inmate_gender <- table(inmates$`Sex:`)
names(inmate_gender) = c('Female', 'Male')

# Display table 
inmate_gender %>%
  kable(caption='Sex Counts')

# Create a pie chart with labels and colors, and show percentages and counts
pie(inmate_gender, labels = paste0(names(inmate_gender), " (", inmate_gender, ")"), 
    col = c("#FF7F50", "#6495ED"), main = "Gender Distribution of Inmates", 
    cex = 0.8, clockwise = TRUE, density = 10)

# Add a legend with percentages and counts
legend("topright", c(paste0(names(inmate_gender), " (", round(100*inmate_gender/sum(inmate_gender), 1), "%, ", inmate_gender, ")")), cex = 0.8, bty = "n")


```
\newpage

## Age

Age distribution of inmates. Note, there are no inmates younger than 18 years old in 
the system since these records are just for the adult facility.

```{r inmate age, echo=FALSE}
# Add a new column to indicate age group by decade
inmates$age_group <- cut(inmates$`Age:`, breaks = seq(0, 100, by = 10), include.lowest = TRUE)

# Count the number of inmates in each age group
counts <- table(inmates$age_group)

# Display table of counts
counts %>%
  kable(caption='Age Range Groupings')

# Create a bar chart with age group distribution
plot <- barplot(counts, main = "Age Group Distribution", xlab = "Age Group", ylab = "Count", col = rainbow(length(counts)))

# Add a labeled line at age 18
#abline(v = 18, col = "red")
#text(x = 18, y = max(counts), label = "18 years old", pos = 1)

```
\newpage

## Race

Breakdown of inmates by race. It is probably safe to assume that the 'W' category label
is probably 'White', and similar, 'A' and 'B' likely could be for 'Asian' and 'Black' 
respectively. The other are too obscure to guess at this point.

Also note that as of the 2020 Census, the overall percentage of black people in 
Lane County is 2.6% while the inmate population is 6.7%, which is a disparity of 
258% relative to County's racial demographic.

```{r inmate race, echo=FALSE, warning=FALSE}
# load necessary libraries
library(ggplot2)

# Count inmate race
inmate_counts <- inmates %>% 
  group_by(`Race:`) %>%
  summarize(count = n())

# Calc percentage
inmate_counts$percentage <- paste0(round(inmate_counts$count / sum(inmate_counts$count) * 100, 1), '%')

# Show table 
inmate_counts %>%
  kable(caption='Inmate Races')

# create the pie chart
ggplot(inmate_counts, aes(x="", y=count, fill=`Race:`)) +
  geom_bar(stat="identity", width=1) +
  coord_polar(theta = "y") +
  labs(title = "Inmate Race Ratios", fill = "Race", x = NULL, y = NULL) +
  theme_void() +
  theme(legend.position = c(1.0, 0.8))

```

\newpage

## Bail

An over look of how many inmates are being held without bail and how many have
bail requirements. This shows the proportion of inmates that have cash or bond 
requirements for bail. The people held with bail represent inmates that are being
held because of money.

```{r inmates bail and measure, echo = FALSE, warning=FALSE}

# Parse bail and bond values and add total bail column
bail_amounts <- charges %>%
  mutate(cash = as.integer(gsub("[\\$,]", "", `Req. Cash Amt:`)), 
         bond = as.integer( gsub("[\\$,]", "", `Req. Bond Amt:`))) %>%
  select(`full name`, cash, bond) %>%
  group_by(`full name`) %>%
  summarize(tot_cash = sum(cash), 
            tot_bond = sum(bond),
            total_bail = sum(cash) + sum(bond))

bail_amounts$bail <- 'No Bail'
bail_amounts$bail[bail_amounts$total_bail > 0] <- 'Bail'

# Create formatted table of bail counts
bail_amounts %>%
  group_by(bail) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  mutate(percentage = paste0(round(percentage, 0), '%')) %>%
  kable(caption="Inmates with/without Bail Requirements")

# Count the number of inmates with each bail type
bail_counts <- table(bail_amounts$bail)

# Create a pie chart to show the proportion of inmates with different bail types
pie(bail_counts, labels = names(bail_counts), col = c('red', "green"))
title(main = "Proportion of Inmates by Bail Type")

```

\newpage

# Charges

## Measure 11

One of the metrics the sheriff has used is concerning inmates awaiting Measure 11 charges. 
Measure 11 defines minimum mandatory sentences for listed felonies. 


> "No pre-trial adults in custody being held on Measure 11 charges have been
> released due to capacity in nearly 10 years (since July 1, 2013)."
> `r tufte::quote_footer('--- Sheriff Cliff Harrold to Board of County Commissioners, on Jan 25, 2023')`

So let us inspect the porportion of inmates that are currently booked on Measure 11 charges.

```{r measure11 stats, echo=FALSE, warning=FALSE}

# Load list of measure 11 codes
measure11 <- read_csv('measure_11.csv', show_col_types = FALSE)

# encode TRUE/FALSE for measure 11 violations
charges$m11 <- FALSE
charges$m11[charges$v_code %in% measure11$code] <- TRUE

# Filter for only m11 charges then get distinct booking info
m11_booking <- charges %>%
  filter(m11 == TRUE) %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# add m11 column to inmate data frame
inmates$measure11 <- FALSE 
inmates$measure11[inmates$`Booking Number:` %in% m11_booking$`Booking Number:`] <- TRUE

# Count the number of inmates labeled as measure 11 and not
measure11_counts <- table(inmates$measure11)

data.frame(measure11_counts) %>%
  kable(caption="Inmates with Measure 11 Charges")

# Create a pie chart with percentage labels
pie(measure11_counts, 
    labels = paste0(names(measure11_counts), ": ", 
                    round(100*measure11_counts/sum(measure11_counts), 2), "%"), 
    col = c("orange", "lightblue"),
    main = "Proportion of inmates with measure 11 charges")

# Add number annotations
legend("topright", legend = names(measure11_counts), 
       title = "Measure 11", 
       fill = c("orange", "lightblue"))

```




```{r measure11 inmates, echo=FALSE, warning=FALSE}

# Filter for only m11 charges then get distinct booking info
m11_booking <- charges %>%
  filter(m11 == TRUE) %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# Order and display inmates
m11_booking %>%
  arrange(`full name`) %>%
  kable(caption='Inmates with Measure 11 Charges')


# charges %>%
#   filter(m11 == TRUE) %>%
#   group_by(`full name`, `v_code`) %>%
#   ungroup() %>%
#   select(`full name`, `Violation:`) %>%
#   arrange(`full name`) %>%
#   kable(caption='Inmates with Measure 11 charges')

```

\newpage

## Top Charges

The majority of inmates are booked for multiple charges. This is a breakdown of 
the total volume of charges on the books. Keep in mind that there are many more 
charges than number of inmates. 

```{r violations, echo=FALSE, warning=FALSE}

# Count all Violations
charge_counts <- charges %>%
  group_by(v_code) %>%
  summarize(count = n()) %>%
  arrange(desc(`count`))

# make list of codes to names
vcode_names <- charges %>%
  select(v_code, `Violation:`) %>%
   distinct(v_code, .keep_all = TRUE)

# merge violation names to violation counts
charge_counts <- merge(x=charge_counts, y=vcode_names,
                    by.x='v_code', by.y='v_code',
                    all.x=TRUE, all.y=FALSE)

# sort and display table of top violations
charge_counts %>%
  select(`Violation:`, count) %>%
  arrange(desc(`count`)) %>%
  head(20) %>%
  kable()


```
\newpage

## Top Non-Measure 11 charges

These are the counts of charges for inmates that are not booked with Measure 11 charges.
That means if someone is booked with a Measure 11 charge along with lesser charges, 
those lesser charges will not be represented here. These are just the charges 
for inmates who could potentially be released due to not being booked with a 
Measure 11 charge.

```{r violation non m11, echo=FALSE, warning=FALSE}

# get charges for people who are not booked with measure 11 charges
non_m11_charges <- charges[!charges$`Booking Number:` %in% m11_booking$`Booking Number:`,]

# sort and display table of top violations
non_m11_charges %>%
  group_by(`Violation:`) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  select(`Violation:`, count) %>%
  arrange(desc(`count`)) %>%
  head(20) %>%
  kable()


```

\newpage



## Non-Measure 11 and Bail

These inmates represent those being held that do not have Measure 11 charges
and are being held with Cash or Bond requirements for bail.

```{r inmates bail, echo = FALSE, warning=FALSE}
# get charges for people who are not booked with measure 11 charges
non_m11_charges <- charges[!charges$`Booking Number:` %in% m11_booking$`Booking Number:`,]

# Parse bail and bond values and add total bail column
bail_amounts <- non_m11_charges %>%
  mutate(cash = as.integer(gsub("[\\$,]", "", `Req. Cash Amt:`)), 
         bond = as.integer( gsub("[\\$,]", "", `Req. Bond Amt:`))) %>%
  select(`full name`, cash, bond) %>%
  group_by(`full name`) %>%
  summarize(tot_cash = sum(cash), 
            tot_bond = sum(bond),
            total_bail = sum(cash) + sum(bond))

bail_amounts$bail <- 'No Bail'
bail_amounts$bail[bail_amounts$total_bail > 0] <- 'Bail'

# Create formatted table of bail counts
bail_amounts %>%
  group_by(bail) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  mutate(percentage = paste0(round(percentage, 0), '%')) %>%
  kable(caption="Non-Measure 11 Inmates Bail Requirements")

# Count the number of inmates with each bail type
bail_counts <- table(bail_amounts$bail)

# Create a pie chart to show the proportion of inmates with different bail types
pie(bail_counts, labels = names(bail_counts), col = c('red', "green"))
title(main = "Proportion of Non-Measure 11 Inmates by Bail Type")

```


\newpage

# Conclusion

The jail is mostly holding inmates that do not have Measure 11 charges. There are currently 43 inmates that have unmet cash bail requirements that are not being held for Measure 11 felonies. Further, 297 inmates do not have Measure 11 charges. 

# Appendix A

source code available at <https://github.com/russl-corey/Lane_County_Jail>

This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details. You should have received a copy of the GNU General Public License along with this program. If not, see [\<https://www.gnu.org/licenses/\>](https://www.gnu.org/licenses/).
